---
title: "Statistical significance"
author: "Thomas Bøjer Rasmussen"
institute: "Department of Clinical Epidemiology<br>Aarhus University Hospital"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "my-css.css"]
    seal: true # Automatic title slide from YAML
    nature:
      self_contained: false
      ratio: '16:9' # Display ratio (default 4:3)
---      

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  dev = "svglite",
  dev.args = list(bg="transparent"),
  fig.width = 9, 
  fig.height = 4
)

library("svglite")
library("RefManageR")

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

```{r load-refs, include=FALSE}
BibOptions(
  check.entries = FALSE,
  bib.style = "numeric",
  cite.style = "numeric",
  style = "text",
  use.regex = TRUE,
  ignore.case = TRUE,
  hyperlink = FALSE,
  dashed = FALSE
)
my_bib <- ReadBib("./bibliography.bib", check = FALSE)
```


## statistical significant testing

Statistical significance testing based on $p$-values is ubiquitous in research

Dichotomization of results into "statistical significant" and "statistical non-significant" is problematic at best

???

notes 

---

## Alternative approach to using p-values

Focus on CI's instead, describing the point estimate and the lower and upper limits.

---

## Interpretation of $p$-values 

Common incorrect interpretations

What is the correct interpretation?

There is a great article talking about around 20 misinterpretations.
Find it.

---

## $p$-values in descriptive tables

Reviewers often request addition of p-values in "table 1"'s.

Nonsensical. Find good references.

---

## Something on D-values?

---



## References

```{r refs, echo=FALSE, results="asis"}
# Use NoCite to add all references that are not cited to the bibliography
NoCite(my_bib, "*")
PrintBibliography(my_bib)
```
